%!TEX root = main.tex
\chapter{Introduction}

The explosion of the mobile market has led to the proliferation of location-aware mobile devices and a wide range of mobile applications that provide location-based content. Augmented reality (AR) applications, for example, enhance the user's real-life environment with location-specific information. The prevalence and affordability of these mobile devices, such as smartphones and tablets, make them a natural choice as tools to augment education and learning. TaleBlazer is an augmented reality location-based game platform that allows users to create their own games that take place in the real world and play them on their mobile devices. A goal of the TaleBlazer project is to determine the educational impact of location-based games, such as how these games motivate users to learn more. TaleBlazer Analytics is an automated system that allows game designers and researchers to gather and analyze anonymous data about users' behaviors during TaleBlazer games. 

\section{Motivations for TaleBlazer Analytics}
The wide gamut of user-created TaleBlazer games requires a data collection system that is both flexible and useful. TaleBlazer Analytics was developed to allow game designers and researchers to get specific metrics about when and how their games are played. For game designers, these metrics allow them to create more engaging and effective game experiences. For researchers, these metrics provide crucial insight into users' gameplay progress and decision making. Game designers and researchers also need a way to quickly analyze data collected from game sessions.

Existing analytics solutions fail to provide data analytics relevant to TaleBlazer specifically and often do not provide an adequate level of user privacy. Furthermore, the unique nature of TaleBlazer games requires custom analysis of the generated data. As a result, it was necessary to develop TaleBlazer Analytics to meet our needs. 

Over the past year and a half, I have developed an automated data collection system which seamlessly integrates with the existing TaleBlazer app and server architecture, working closely at each step with the TaleBlazer team. TaleBlazer Analytics is comprised of a backend server, mobile client, and web application; these individual components are jointly responsible for the collection, storage, and analysis of TaleBlazer gameplay data. 

\section{Chapter Summary}
This thesis describes the background, design, and development of the TaleBlazer Analytics system. Chapter 2 details TaleBlazer in-depth and expands on the need for TaleBlazer Analytics. Chapter 3 explains the design process and preliminary work that was performed prior to the start of development. Chapter 4 details the TaleBlazer Analytics system and its component in depth. Chapter 5 goes into particular tests that were performed on the system and their effects on the project. Chapter 6 proposes future work for the project and Chapter 7 details the overall contributions of this thesis.




